Modeling Lake Trophic State: A Random Forest Approach

### 2015-04-03 inst/doc/manuscript.Rmd
final clearance version

---
title: 'Modeling Lake Trophic State: A Data Mining Approach'
author: 
- name: Jeffrey W. Hollister
  affilnum: 1
  email: hollister.jeff@epa.gov
- name: W. Bryan Milstead
  affilnum: 1
- name: Betty J. Kreakie
  affilnum: 1
affiliation:
- affilnum: 1
  affil: US Environmental Protection Agency, Office of Research and Development, National     Health and Environmental Effects Research Laboratory, Atlantic Ecology Division, 27 Tarzwell Drive  Narragansett, RI, 02882, USA
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    template: components/manuscript.latex
fontsize: 11pt
capsize: normalsize
csl: components/ecology.csl
documentclass: article
spacing: doublespacing
linenumbers: yes
bibliography: components/LakeTrophicModelling.bib
abstract: no
---

@@

```{r setup, include=FALSE, echo=FALSE}
#devtools::install_github("USEPA/LakeTrophicModelling",quick=TRUE,
#                        auth_token="8742a8c1c2207393659a1de1bd53527fa56ea0d7")
#devtools::install_github("jhollist/wesanderson")
library("wesanderson")
library("LakeTrophicModelling")
library("knitr")
library("ggplot2")
library("sp")
library("rgdal")
library("e1071")
library("dplyr")
library("tidyr")
opts_chunk$set(dev = 'pdf', fig.width=6, fig.height=4.3)
data(LakeTrophicModelling)

#Checks for existing cache (from another project)
if(file.exists("prior_cache")){
  for(i in gsub(".rdb","",list.files("prior_cache",".rdb",full.names=T))){
    lazyLoad(i)
  }
}

@@

```{r analysis , eval=FALSE, include=FALSE, echo=FALSE, cache=TRUE}
################################################################################
#All analysis in here, that way all bits of paper have access to final objects
#Tables, figures, and numerical results placed at end of paper.
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
#This chunk takes several days to complete - DO NOT EDIT!!!
################################################################################
#Model 1: All Variables, 4 Trophic State Classes
  #Variable Selction
  all_ts4<-iterVarSelRF(ltmData[predictors_all],ltmData$TS_CHLA_4,numCore=4,100,
        								ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
  											vars.drop.num=1,outStr="all_ts4",time=TRUE)
  #Random Forest
  all_ts4_vars<-unique(unlist(all_ts4[1:100]))
  all_ts4_rf<-runRandomForest(ltmData[all_ts4_vars],ltmData$TS_CHLA_4, 
                              ntree=10000,importance=TRUE,proximity=TRUE)
################################################################################
#Model 2: All Variables, 3 Trophic State Classes
  #Variable Selction
  all_ts3<-iterVarSelRF(ltmData[predictors_all],ltmData$TS_CHLA_3,numCore=4,100,
      									ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
  											vars.drop.num=1,outStr="all_ts3",time=TRUE)
  #Random Forest
  all_ts3_vars<-unique(unlist(all_ts3[1:100]))
  all_ts3_rf<-runRandomForest(ltmData[all_ts3_vars],ltmData$TS_CHLA_3, 
                              ntree=10000,importance=TRUE,proximity=TRUE)
################################################################################
#Model 3: All Variables, 2 Trophic State Classes
  #Variable Selction
  all_ts2<-iterVarSelRF(ltmData[predictors_all],ltmData$TS_CHLA_2,numCore=4,100,
      									ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
  											vars.drop.num=1,outStr="all_ts2",time=TRUE)

  #Random Forest
  all_ts2_vars<-unique(unlist(all_ts2[1:100]))
  all_ts2_rf<-runRandomForest(ltmData[all_ts2_vars],ltmData$TS_CHLA_2, 
                              ntree=10000,importance=TRUE,proximity=TRUE)
################################################################################
#Model 4: GIS Only Variables, 4 Trophic State Classes
  #Variable Selction
  gis_ts4<-iterVarSelRF(ltmData[predictors_gis],ltmData$TS_CHLA_4,numCore=4,100,
          							ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
  											vars.drop.num=1,outStr="gis_ts4",time=TRUE)

  #Random Forest
  gis_ts4_vars<-unique(unlist(gis_ts4[1:100]))
  gis_ts4_rf<-runRandomForest(ltmData[gis_ts4_vars],ltmData$TS_CHLA_4, 
                              ntree=10000,importance=TRUE,proximity=TRUE)
################################################################################
#Model 5: GIS Only Variables, 3 Trophic State Classes
  #Variable Selction
  gis_ts3<-iterVarSelRF(ltmData[predictors_gis],ltmData$TS_CHLA_3,numCore=4,100,
        								ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
  											vars.drop.num=1,outStr="gis_ts3",time=TRUE)

  #Random Forest
  gis_ts3_vars<-unique(unlist(gis_ts3[1:100]))
  gis_ts3_rf<-runRandomForest(ltmData[gis_ts3_vars],ltmData$TS_CHLA_3, 
                              ntree=10000,importance=TRUE,proximity=TRUE)
################################################################################
#Model 6: GIS Only Variables, 2 Trophic State Classes
  #Variable Selction
  gis_ts2<-iterVarSelRF(ltmData[predictors_gis],ltmData$TS_CHLA_2,numCore=4,100,
      									ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
  											vars.drop.num=1,outStr="gis_ts2",time=TRUE)
  #Random Forest
  gis_ts2_vars<-unique(unlist(gis_ts2[1:100]))
  gis_ts2_rf<-runRandomForest(ltmData[gis_ts2_vars],ltmData$TS_CHLA_2, 
                              ntree=10000,importance=TRUE,proximity=TRUE)

@@

```{r summaryStats, echo=FALSE, cache=TRUE, warning=FALSE}
acc_results<-data.frame(model_number=c("Model 1","Model 2","Model 3","Model 4","Model 5","Model 6"),
                        total_accuracy=c(classAgreement(all_ts4_rf$confusion[,1:4])$diag,
                                         classAgreement(all_ts3_rf$confusion[,1:3])$diag,
                                         classAgreement(all_ts2_rf$confusion[,1:2])$diag,
                                         classAgreement(gis_ts4_rf$confusion[,1:4])$diag,
                                         classAgreement(gis_ts3_rf$confusion[,1:3])$diag,
                                         classAgreement(gis_ts2_rf$confusion[,1:2])$diag),
                        kappa_coeff=c(classAgreement(all_ts4_rf$confusion[,1:4])$kappa,
                                         classAgreement(all_ts3_rf$confusion[,1:3])$kappa,
                                         classAgreement(all_ts2_rf$confusion[,1:2])$kappa,
                                         classAgreement(gis_ts4_rf$confusion[,1:4])$kappa,
                                         classAgreement(gis_ts3_rf$confusion[,1:3])$kappa,
                                         classAgreement(gis_ts2_rf$confusion[,1:2])$kappa))
var_importance<-rbind_list(varImportance(all_ts4_rf,"Model 1"),
                           varImportance(all_ts3_rf,"Model 2"),
                           varImportance(all_ts2_rf,"Model 3"),
                           varImportance(gis_ts4_rf,"Model 4"),
                           varImportance(gis_ts3_rf,"Model 5"),
                           varImportance(gis_ts2_rf,"Model 6"))
var_importance$variable_names<-factor(var_importance$variable_names)
#var_importance$model_id<-factor(var_importance$model_id)
var_importance<-inner_join(var_importance,data_def[c("variable_names","description")],"variable_names")
var_importance<-var_importance[order(var_importance$model_id,-var_importance$mean_decrease_gini, decreasing=FALSE),]

@@

**Abstract**  
Productivity of lentic ecosystems has been well studied and it is widely accepted that as nutrient inputs increase,  productivity increases and lakes transition from low trophic state (e.g. oligotrophic) to higher trophic states  (e.g. eutrophic). These broad trophic state classifications are good predictors of ecosystem health and ecosystem services/disservices (e.g. recreation, aesthetics, fisheries, and harmful algal blooms). While the relationship between  nutrients and trophic state provides reliable predictions, it requires *in situ* water quality data in order to paramterize the model.  This limits the application of these models to lakes with existing and, more importanly, available water quality data.  To expand our ability to predict in lakes without water quality data, we take advantage of the availability of a large national lakes water quality database, land use/land cover data, lake morphometry data, other universally available data, and modern data mining approaches to build and assess models of lake tropic state that may be more universally applied.  We use random forests and random forest variable selection to identify variables to be used for predicting trophic state and we compare the performance of two models of trophic state (as determined by chlorophyll *a* concentration).   The first model estimates trophic state with *in situ* as well as universally available data and the second model uses universally available data only. For each of these models we used three separate trophic state categories, for a total of six models. Overall accuracy for models buils from *in situ* and universal data ranged from `r round(min(acc_results[1:3,2]),3)`% to `r round(max(acc_results[1:3,2]),3)`%.  For the universal data only models, Overall accuraccy ranged from `r round(min(acc_results[4:6,2]),3)`% to `r round(max(acc_results[4:6,2]),3)`%. Lastly, it is believed that the presence and abundance of cyanobacteria is strongly associated with trophic state.  To test this we examine the association between estimates of cyanobacteria biovolume and the measured and predicted trophic state and find a positive relationship.  Expanding these preliminary results to include cyanobacteria taxa indicates that cyanobacteria are significantly more likely to be found in highly eutrophic lakes. These results suggest that predictive models of lake trophic state may be improved with additional information on the landscape surrounding lakes and that those models provide additional information on the presence of potentially harmful cyanobacteria taxa.

@@

#Introduction
Productivity in lentic systems is often categorized across a range of tropic states (e.g. the tropic continuum) from early succesional (i.e. oligotrophic)to late successional lakes (i.e. hypereutrophic) [@carlson1977trophic] and lakes naturally occur across this range. Naturally oligotrophic lakes occur in nutrient poor areas or have a more recent geologic history.  These lakes are often found in higher elevations, have clear water, and are often favored for drinking water or direct contact recreation (e.g. swimming).   Lakes with higher productivity (e.g. eutrophic lakes) have greater nutrient loads, tend to be less clear, have greater density of aquatic plants, and often support more diverse and abundant fish communities. Higher primary productivity is not necessarily a predictor of poor ecological condition.

It is natural for lakes to shift from lower to higher trophic states but this is a slow process.  Given this fact, monitoring trophic state allows the identification of rapid shifts in trophic state or locating lakes with unusually high productivity (e.g. hypereutrophic).  These cases are indicative of lakes under greater anthropogenic nutrient loads, also known as cultural eutrophication, and are more likely to be at risk of fish kills, fouling, and harmful algal blooms[@smith1998cultural;@smith1999eutrophication;@smith2006eutrophication].  Given the association between trophic state and many ecosystem services and disservices, being able to model trophic state could allow for estimating trophic state in unmonitored lakes and provide a first cut at identifying lakes with the potential for harmful algal blooms and other problems associated with cultural eutrophication.

As trophic state can be defined by a number of *in situ* water quality measurements, most models have used this information as predictors.  This leads to accurate models, but also requires data that is sparse and not always availble, limiting the population of lakes that can be modeled.  Landscape data is ubiqutious ...

We have three goals for this preliminary research.  First, we build and assess multiple models of lake trophic state using a full suite of data including *in situ* water quality and universally available data (e.g. landscape data). Second, we assess the accuracy of predicted trophic state in lakes with only the universally available data.  Lastly, we explore associations between trophic state and cyanobacteria to explore.

@@

##Data and Study Area
We utilize four primary sources of data for this study,the National Lakes Assessment (NLA), the National Lake Cover Dataset (NLCD), modeled lake morphometery,  and estimated cyanobacteria biovolumes [@usepa2009national;@homer2004development;@xian2009updating;@hollister2010volume;@hollister_predicting_2011;@lakemorpho2014;@beaulieu2013nutrients].  All datasets are national in scale and provide a unique snapshot view of the condition of lakes in the United States'.

The NLA data were collected during the summer of 2007 and the final data were released in 2009.  With consistent methods and metrics collected at 1056 locations across the conterminous United States (Figure \ref{fig:nlaMap}), the NLA provides a unique opportunity to examine broad scale patterns in lake productivity.  The NLA collected data on biophysical measures of lake water quality and habitat.  For this analysis we primarily examined the water quality measurements from the NLA [@usepa2009national].  Adding to the monitoring data collected via the NLA, we use the 2006 NLCD data to examine the possible landscape-level drivers of trophic status in lakes.  The NLCD is a nationally collected land use land cover dataset that also provides estimates of impervious surface.  We collected total land use land cover and total percent impervious surface within a 3 kilometer buffer surrounding the lake to examine larger landscape-level effect [@homer2004development;@xian2009updating]. We also used various measures of lake morphometry (i.e. depth, volume, fetch, etc.) as they are important in understanding lake productivity, yet many of these data are difficult to obtain for large numbers of lakes over broad regions.  To add this information we modeled lake morphometry [@hollister2010volume;@hollister_predicting_2011;@lakemorpho2014;@NLMDInPrep]. Lastly, to explore associations between trophic state and cyanobacteria, we used estimates of cyanobacterial biovoulme caluclated by Beaulieu *et al.* [-@beaulieu2013nutrients].  Cyanobacteria biovolumes are a truer measure of cyanobacteria dominance than abundance as there is great variability in the size within and between species.  We have consolidated the taxa level estimates from Beaulieu *et al.* [-@beaulieu2013nutrients] and  summed that information on a per-lake basis.

##Predicting Trophic State with Random Forests
Random forest is a machine learning algorithm that aggregates numerous decision trees in order to obtain a consensus prediction of the response categories [@breiman2001random].  Bootstrapped sample data is recursively partitioned according to a given random subset of predictor variables and completely grown without pruning.   With each new tree, both the sample data and predictor variable subset is randomly selected.

While random forests are able to handle numerous correlated variables without a decrease in prediction accuracy, unusually large numbers of related variables can reduce accuracy and increase the chances of over-fitting the model.  This is a problem often faced in gene selection and in that field, a variable selection method based on random forest has been succesfully applied [@diaz2006gene].  We use varselRF in R to initially examine the importance of the water quality and GIS derived variables and select a subset, the reduced model, to then pass to random forest[@diaz-uriarte2010varSelRF].

Using R's randomForest package, we pass the reduced models selected with varSelRF and calculate confusion matrices, overall accuracy and kappa coeffecient [@liaw2002randomForest]. From the reduced model random forests we collect a consensus prediction and calculate a confusion matrix and summary stats.

##Model Details
Using a combination of the `varSelRF` and `randomForest` we ran models for six combinations of variables and trophic state classifications.  These combinations included different combinations of the Chlorphyll *a* trophic states (Table \ref{tab:trophicStateTable}) along with all variables and the GIS only variables (i.e. no *in situ* infromation).  The six model combinations were:

1. Chlorophyll *a* trophic state - 4 class = All variables (*in situ* water quality, lake morphometry, and landscape)
2. Chlorophyll *a* trophic state - 3 class = All variables (*in situ* water quality, lake morphometry, and landscape)
3. Chlorophyll *a* trophic state - 2 class = All variables (*in situ* water quality, lake morphometry, and landscape)
4. Chlorophyll *a* trophic state - 4 class = All variables (lake morphometry, and landscape)
5. Chlorophyll *a* trophic state - 3 class = All variables (lake morphometry, and landscape)
6. Chlorophyll *a* trophic state - 2 class = All variables (lake morphometry, and landscape)

#Results and Discussion

>>>>

---
title: 'Modeling Lake Trophic State: A Random Forest Approach'
author:
- affilnum: 1
  email: hollister.jeff@epa.gov
  name: Jeffrey W. Hollister
- affilnum: 1
  name: W. Bryan Milstead
- affilnum: 1
  name: Betty J. Kreakie
affiliation:
- affil: US Environmental Protection Agency, Office of Research and Development, National     Health
    and Environmental Effects Research Laboratory, Atlantic Ecology Division, 27 Tarzwell
    Drive  Narragansett, RI, 02882, USA
  affilnum: 1
output:
  pdf_document: 
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    template: components/manuscript.latex
  word_document: yes
capsize: normalsize
csl: components/freshwater-science.csl
documentclass: article
fontsize: 11pt
linenumbers: yes
bibliography: components/LakeTrophicModelling.bib
spacing: doublespacing
abstract: no
---

@@

```{r setup, include=FALSE, echo=FALSE}
if(!require(LakeTrophicModelling)){
  devtools::install_github("USEPA/LakeTrophicModelling",quick=TRUE)
}
if(!require(wesanderson)){
devtools::install_github("jhollist/wesanderson")
}
if(!require(condprob2)){
devtools::install_github("jhollist/condprob2")
}
if(!require(tidyr)){
  install.packages("tidyr")
}
library("wesanderson")
library("LakeTrophicModelling")
library("knitr")
library("ggplot2")
library("sp")
library("rgdal")
library("e1071")
library("dplyr")
library("tidyr")
library("condprob2")
opts_chunk$set(dev = 'png', fig.width=7.5)
data(LakeTrophicModelling)

#Checks for existing cache (from another project)
#Else if Repeats running chunk outside of knitr
if(file.exists("prior_cache")){
  for(i in gsub(".rdb","",list.files("prior_cache",".rdb",full.names=T))){
    lazyLoad(i)
  }
} else if(file.exists("vignettes/prior_cache")){
  for(i in gsub(".rdb","",list.files("vignettes/prior_cache",".rdb",full.names=T))){
    lazyLoad(i)
  }
}

@@

```{r analysis , eval=FALSE, include=FALSE, echo=FALSE, cache=FALSE}
################################################################################
#All analysis in here, that way all bits of paper have access to final objects
#Tables, figures, and numerical results placed at end of paper.
#Included for completeness.  Results already run in another project
#Those results loaded via lazyLoad in setup chunk
#This chunk takes several days to complete - DO NOT EDIT!!!
################################################################################
#Model 1: All Variables, 4 Trophic State Classes
  #Variable Selction
  #all_ts4<-iterVarSelRF(ltmData[predictors_all],ltmData$TS_CHLA_4,numCore=4,100,
  #      								ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
  #  											vars.drop.num=1,outStr="all_ts4",time=TRUE)
  #all_ts4_rf_full<-runRandomForest(ltmData[predictors_all],ltmData$TS_CHLA_4, 
  #                          ntree=10000,importance=TRUE,proximity=TRUE)
  #Random Forest
  all_ts4_vars<-unique(unlist(all_ts4[1:100]))
  all_ts4_rf<-runRandomForest(ltmData[all_ts4_vars],ltmData$TS_CHLA_4, 
                              ntree=10000,importance=TRUE,proximity=TRUE)
################################################################################
#Model 2: All Variables, 3 Trophic State Classes
  #Variable Selction
  #all_ts3<-iterVarSelRF(ltmData[predictors_all],ltmData$TS_CHLA_3,numCore=4,100,
  #    									ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
  #  											vars.drop.num=1,outStr="all_ts3",time=TRUE)
  # all_ts3_rf_full<-runRandomForest(ltmData[predictors_all],ltmData$TS_CHLA_3, 
  #                            ntree=10000,importance=TRUE,proximity=TRUE)
  #Random Forest
  all_ts3_vars<-unique(unlist(all_ts3[1:100]))
  all_ts3_rf<-runRandomForest(ltmData[all_ts3_vars],ltmData$TS_CHLA_3, 
                              ntree=10000,importance=TRUE,proximity=TRUE)
################################################################################
#Model 3: All Variables, 2 Trophic State Classes
  #Variable Selction
  #all_ts2<-iterVarSelRF(ltmData[predictors_all],ltmData$TS_CHLA_2,numCore=4,100,
  #    									ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
  #											vars.drop.num=1,outStr="all_ts2",time=TRUE)
  #all_ts2_rf_full<-runRandomForest(ltmData[predictors_all],ltmData$TS_CHLA_2, 
  #                           ntree=10000,importance=TRUE,proximity=TRUE)
  #Random Forest
  all_ts2_vars<-unique(unlist(all_ts2[1:100]))
  all_ts2_rf<-runRandomForest(ltmData[all_ts2_vars],ltmData$TS_CHLA_2, 
                              ntree=10000,importance=TRUE,proximity=TRUE)
################################################################################
#Model 4: GIS Only Variables, 4 Trophic State Classes
  #Variable Selction
  #gis_ts4<-iterVarSelRF(ltmData[predictors_gis],ltmData$TS_CHLA_4,numCore=4,100,
  #        							ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
  #											vars.drop.num=1,outStr="gis_ts4",time=TRUE)
  #gis_ts4_rf_full<-runRandomForest(ltmData[predictors_gis],ltmData$TS_CHLA_4, 
  #                            ntree=10000,importance=TRUE,proximity=TRUE)
  #Random Forest
  gis_ts4_vars<-unique(unlist(gis_ts4[1:100]))
  gis_ts4_rf<-runRandomForest(ltmData[gis_ts4_vars],ltmData$TS_CHLA_4, 
                              ntree=10000,importance=TRUE,proximity=TRUE)
################################################################################
#Model 5: GIS Only Variables, 3 Trophic State Classes
  #Variable Selction
  #gis_ts3<-iterVarSelRF(ltmData[predictors_gis],ltmData$TS_CHLA_3,numCore=4,100,
  #      								ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
  #											vars.drop.num=1,outStr="gis_ts3",time=TRUE)
  #gis_ts3_rf_full<-runRandomForest(ltmData[predictors_gis],ltmData$TS_CHLA_3, 
  #                            ntree=10000,importance=TRUE,proximity=TRUE)
  #Random Forest
  gis_ts3_vars<-unique(unlist(gis_ts3[1:100]))
  gis_ts3_rf<-runRandomForest(ltmData[gis_ts3_vars],ltmData$TS_CHLA_3, 
                              ntree=10000,importance=TRUE,proximity=TRUE)
################################################################################
#Model 6: GIS Only Variables, 2 Trophic State Classes
  #Variable Selction
  #gis_ts2<-iterVarSelRF(ltmData[predictors_gis],ltmData$TS_CHLA_2,numCore=4,100,
  #    									ntree=10000,ntreeIterat=5000,vars.drop.frac=NULL,
  #											vars.drop.num=1,outStr="gis_ts2",time=TRUE)
  #gis_ts2_rf_full<-runRandomForest(ltmData[predictors_gis],ltmData$TS_CHLA_2, 
  #                            ntree=10000,importance=TRUE,proximity=TRUE)
  #Random Forest
  gis_ts2_vars<-unique(unlist(gis_ts2[1:100]))
  gis_ts2_rf<-runRandomForest(ltmData[gis_ts2_vars],ltmData$TS_CHLA_2, 
                              ntree=10000,importance=TRUE,proximity=TRUE)

@@

```{r summaryStats, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE}
acc_results<-data.frame(model_number=c("Model 1","Model 2","Model 3","Model 4","Model 5","Model 6"),
                        total_accuracy=c(classAgreement(all_ts4_rf$confusion[,1:4])$diag,
                                         classAgreement(all_ts3_rf$confusion[,1:3])$diag,
                                         classAgreement(all_ts2_rf$confusion[,1:2])$diag,
                                         classAgreement(gis_ts4_rf$confusion[,1:4])$diag,
                                         classAgreement(gis_ts3_rf$confusion[,1:3])$diag,
                                         classAgreement(gis_ts2_rf$confusion[,1:2])$diag),
                        kappa_coeff=c(classAgreement(all_ts4_rf$confusion[,1:4])$kappa,
                                         classAgreement(all_ts3_rf$confusion[,1:3])$kappa,
                                         classAgreement(all_ts2_rf$confusion[,1:2])$kappa,
                                         classAgreement(gis_ts4_rf$confusion[,1:4])$kappa,
                                         classAgreement(gis_ts3_rf$confusion[,1:3])$kappa,
                                         classAgreement(gis_ts2_rf$confusion[,1:2])$kappa))
var_importance<-rbind(varImportance(all_ts4_rf,"Model 1"),
                           varImportance(all_ts3_rf,"Model 2"),
                           varImportance(all_ts2_rf,"Model 3"),
                           varImportance(gis_ts4_rf,"Model 4"),
                           varImportance(gis_ts3_rf,"Model 5"),
                           varImportance(gis_ts2_rf,"Model 6"))
var_importance$variable_names<-factor(var_importance$variable_names)
#var_importance$model_id<-factor(var_importance$model_id)
var_importance<-inner_join(var_importance,data_def[c("variable_names","description")],"variable_names")
var_importance<-var_importance[order(var_importance$model_id,-var_importance$mean_decrease_gini, decreasing=FALSE),]

#Example Votes
lakeVotes<-data.frame(all_ts4_rf$y,all_ts4_rf$predicted,all_ts4_rf$votes)
lakeVotes<-data.frame(lakeVotes,max_prob=apply(lakeVotes[,3:6],1,max))
lakeVotes$all_ts4_rf.predicted<-factor(lakeVotes$all_ts4_rf.predicted,ordered=T)
lakeVotes<-data.frame(lakeVotes,correct_binary=lakeVotes$all_ts4_rf.y==lakeVotes$all_ts4_rf.predicted)
lakeVotes<-data.frame(COMID=getLakeIDs(ltmData[all_ts4_vars],ltmData$TS_CHLA_4,ltmData$Comid),lakeVotes)
lakeVotes<-na.omit(lakeVotes)
```

@@

**Abstract**  
Productivity of lentic ecosystems has been well studied and it is widely accepted that as nutrient inputs increase,  productivity increases and lakes transition from lower trophic state (e.g. oligotrophic) to higher trophic states  (e.g. eutrophic). These broad trophic state classifications are good predictors of ecosystem condition and ecosystem services and disservices (e.g. recreation, aesthetics, fisheries, and harmful algal blooms). While the relationship between  nutrients and trophic state provides reliable predictions, it requires *in situ* water quality data in order to parameterize the model.  This limits the application of these models to lakes with existing and, more importantly, available water quality data.  To expand our ability to predict trophic state in lakes without water quality data, we take advantage of the availability of a large national lakes water quality database (i.e. the National Lakes Assessment), land use/land cover data, lake morphometry data, other universally available data, and modern data mining approaches to build and assess models of lake trophic state that may be more universally applied.  We use random forests and random forest variable selection to identify variables to be used for predicting trophic state and we compare the performance of two sets of models of trophic state (as determined by chlorophyll *a* concentration).   The first set of models estimates three different trophic state classification with *in situ* as well as universally available data and the second set of models uses universally available GIS data only. Overall accuracy for models built from *in situ* and universal data ranged from `r round(min(acc_results[1:3,2]),2)*100`% to `r round(max(acc_results[1:3,2]),2)*100`%.  For the universal data only models, overall accuracy ranged from `r round(min(acc_results[4:6,2]),2)*100`% to `r round(max(acc_results[4:6,2]),2)*100`%. Lastly, presence and abundance of cyanobacteria is often strongly associated with trophic state.  To test this we examine the association between estimates of cyanobacteria abundance and measured chlorophyll *a* and find a positive relationship.  These results suggest that predictive models of lake trophic state may be improved with additional information on the landscape surrounding lakes and that those models provide additional information on the presence of potentially harmful cyanobacteria taxa. Lastly, the source code and data for this manuscript are freely available from https://github.com/USEPA/LakeTrophicModelling.

@@

#Introduction
Productivity in lentic systems is often categorized across a range of trophic states (e.g. the trophic continuum) from early successional (i.e. oligotrophic) to late successional lakes (i.e. hypereutrophic) with lakes naturally occurring across this range [@carlson1977trophic]. Oligotrophic lakes occur in nutrient poor areas or have a more recent geologic history, are often found in higher elevations, have clear water, and are usually favored for drinking water or direct contact recreation (e.g. swimming).   Lakes with higher productivity (e.g. mesotrophic and eutrophic lakes) have greater nutrient loads, tend to be less clear, have greater density of aquatic plants, and often support more diverse and abundant fish communities. Higher primary productivity is not necessarily a predictor of poor ecological condition as it is natural for lakes to shift from lower to higher trophic states but this is a slow process. However, at the highest productivity levels (hypereutrophic lakes) biological integrity is compromised [@hasler1969cultural; @schindler2008algal; @smith1999eutrophication].

Monitoring trophic state allows the identification of rapid shifts in trophic state or locating lakes with unusually high productivity (e.g. hypereutrophic).  These cases are indicative of lakes under greater anthropogenic nutrient loads, also known as cultural eutrophication, and are more likely to be at risk of fish kills, fouling, and harmful algal blooms [@smith1998cultural;@smith1999eutrophication;@smith2006eutrophication].  Given the association between trophic state and many ecosystem services and disservices, being able to accurately model trophic state could provide a first cut at identifying lakes with the potential for harmful algal blooms (i.e. from cyanobacteria) or other problems associated with cultural eutrophication.  This type of information could be used for setting priorities for managment and allow for more effecient use of limited resources.

As trophic state and related indices can be best defined by a number of *in situ* water quality parameters (modeled or measured), most models have used this information as predictors [e.g., @carvalho_cyanobacterial_2011;@milstead2013estimating;@imboden1978dynamic;@salas1991simplified].  This leads to accurate models, but also requires data that are often sparse and not always available, thus limiting the population of lakes for which we can make predictions.  A possible solution for this is to build models that use widely available data that are correlated to many of the *in situ* variables. For instance, landscape metrics of forests, agriculture, wetlands, and urban land in contributing watersheds have all been shown to explain a significant proportion of the variation (ranging from 50-86%, depending on study) in nutrients in receiving waters [@jones2001predicting; @jones2004importance; @seilheimer2013landscape].  Building on these previously identified associations might allow us to use only landscape and other universally available data to build models.  Identifying predictors using this type of ubiquitous data would allow for estimating trophic state in both monitored and unmonitored lakes.

Many published models of nutrients and trophic state in freshwater systems are based on linear modelling methods such as standard least squares regression or linear mixed models [e.g., @jones2004importance; @jones2001predicting].  While these methods have proven to be reliable, they have limitations (e.g. independence and distribution assumptions, and outlier sensitivity).  Using data mining approaches, such as random forests, avoids many of the limitations, may reduce bias and often provides better predictions [@cutler2007random; @peters2007random; @breiman2001random].  For instance, random forests are non-parametric and thus the data do not need to come from a specific distribution (e.g. Gaussian) and can contain collinear variables [@cutler2007random]. Second, random forests work well with very large numbers of predictors [@cutler2007random].  Lastly, random forests can deal with model selection uncertainty as predictions are based upon a consensus of many models and not just a single model selected with some measure of goodness of fit.

To build on past work, we have identified five areas in which our research contributes.  First, we update trophic state modelling efforts with the use of random forests.  Second, we assess the accuracy of predicted trophic state in lakes with the a) *in situ* and universally available GIS data  and then b) with the universally available GIS data only. Third, we identify important variables for describing lake trophic state.  Fourth, we explore associations between trophic state and cyanobacteria, a key taxa in harmful algal blooms, so that we may begin to understand how changes in trophic state may be linked to an important ecosystem disservice.  Lastly, a key goal of this research is to provide full access to the data and code we used in this research.  As such, this paper, the code, and the data used in the models is made available as an R package from https://github.com/USEPA/LakeTrophicModeling.

@@

##Data and Study Area
We utilize three primary sources of data for this study, the National Lakes Assessment (NLA), the National Land Cover Dataset (NLCD), and  lake morphometery modeled from the NHDPlus and National Elevation Data Set [@usepa2009national;@homer2004development;@xian2009updating;@hollister2010volume;@hollister_predicting_2011;@lakemorpho2014].  All datasets are national in scale and provide a unique snapshot view of the condition of lakes in the conterminous United States during the summer of 2007.

The NLA data were collected during the summer of 2007 and the final data were released in 2009 [@usepa2009national].  With consistent methods and metrics collected at 1056 locations across the conterminous United States (Figure \ref{fig:nlaMap}), the NLA provides a unique opportunity to examine broad scale patterns in lake productivity.  The NLA collected data on biophysical measures of lake water quality and habitat as well as an assessment of the phytoplankton community.  For this analysis we examined the water quality measurements and total cyanobacteria abundance from the National Lakes Assessment [@usepa2009national].

Adding to the monitoring data collected via the NLA, we use the 2006 NLCD data to examine landscape-level drivers of trophic status in lakes.  The NLCD is a national land use/land cover dataset that also provides estimates of impervious surface.  We calculated total proportion of each NLCD land use land cover class and total percent impervious surface within a 3 kilometer buffer surrounding each lake [@homer2004development;@xian2009updating].  A three kilometer buffer was selected as an intermediate measure of the adjacent neighborhood; the three kilometer buffer size is greater than the immediate parcel but smaller than regional  and whole-basin measures.

To account for unique aspects of each lake and characterize lake productivity, we also used measures of lake morphometry (i.e. depth, volume, fetch, etc.).  As these data are difficult to obtain for large numbers of lakes over broad regions, we used modeled estimates of lake morphometry [@hollister2010volume;@hollister_predicting_2011;@lakemorpho2014]. These included: surface area, shoreline length, Shoreline Development, Maximum Depth, Mean Depth, Lake Volume, Maximum Lake Length, Mean Lake Width, Maximum Lake Width, and Fetch.

##Predicting Trophic State with Random Forests
Random forest is a machine learning algorithm that aggregates numerous decision trees in order to obtain a consensus prediction of the response categories [@breiman2001random].  Bootstrapped sample data are recursively partitioned according to a given random subset of predictor variables and a predetermined number of decision trees  are developed.   With each new tree, the sample data subset is randomly selected and with each new split, the subset of predictor variables are randomly selected.  A detailed discussion of the benefits of a random forest approach is beyond the scope of this paper.  To find out more see Breiman [-@breiman2001random] and Cutler et al. [-@cutler2007random].

Random forests are able to handle numerous correlated variables without a decrease in prediction accuracy; however, one possible downfall to this approach is that the resulting model may be difficult to interpret.  This is a problem often faced in gene selection and in that field, a variable selection method based on random forest has been successfully applied and implemented in the R Language as the `varSelRF` package [@diaz2006gene].  With this method, a minimum set of variables that maximizes model accuracy is provided.  This allows us to start with a full suite of predictor variables from which to select a minimum, easier to interpret set of variables.  One issue with the approach in `varSelRF` is that because of the randomization inherent in random forests it is possible to get variation in the minimum selected set of variables.  To account for this we repeated `varSelRF` 100 times.  In our case, repeating the procedure 100 times quickly converged on a set of all possible important variables.

##Model Details
Using both the `varSelRF` and `randomForest` R packages we ran models for six sets of variables and trophic state classifications [@diaz-uriarte2010varSelRF; @liaw2002randomForest].  These included three different combinations of the chlorophyll *a* trophic states as the dependent variables and using all variables (*in situ* and GIS variables) or the GIS only variables (i.e. no *in situ* information) as the independent variables in the random forest.  A listing of all considered variables is in Appendix 1.  Trophic state was defined using the NLA chlorophyll *a* trophic state cut offs and the three combinations of trophic state were used to highlight the possible error caused by misclassification of adjacent classes, such as mesotrophic and eutrophic (Table \ref{tab:trophicStateTable}). Our separation of predictors was chosen so that we could highlight the additional predictive performance provided by adding the *in situ* water quality variables on top of the GIS only variables.  Lastly, we used only complete cases (i.e. missing data were removed) so the total number of observations varied between models.

The six model combinations were:
